Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-06 rnni_convexity.tex
link to implementations in Algorithms section
minor working changes

The algorithm $\findpath$ can be used to approximate the $\rnni$ distance.
We will study the accuracy of this approximation by computing the exact $\rnni$ distance for small trees.
Specifically, we will use the algorithm for computing the complete $\rnni$ graph that is given in \autocite[their Section 3.3]{Gavryushkin2018-ol}.
After pre-computing the graph, we use the standard algorithms~\autocite{Floyd1962-ew,Dijkstra1959-ph} for computing shortest paths.
However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances for larger trees.
Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we compute the $\rnni$ graph on up to six taxa for our analyses.
This computational result will also be used later in the paper to support Conjecture~\ref{conjecture:cluster_theorem}, a modification of Conjecture~9 in \autocite{Gavryushkin2018-ol}, which we prove to be false in this paper.
Our implementations can be found on GitHub in \autocite{Collienne2019}.

@@

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
A path $p$ from $T$ to $R= [x_1, x_2, \ldots, x_n]$ is computed iteratively such that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
Specifically, in step $k$ ($k = 0, \ldots, n-3$) taxon $x_{n-k}$ is moved up to the position it has in $R$.
Here we say that a taxon moves up when it swaps position with its neighboured taxon, which corresponds to an $\nni$ move.
Notice that there might be more than just one such move per step, which means that $p$ can be extended by more than one tree in each step.
After $n-2$ steps we reach $R$.

The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
Our implementation of this algorithm can be found on GitHub \autocite{Collienne2019}.

>>>>

The algorithm $\findpath$ can be used to approximate the $\rnni$ distance.
We will study the accuracy of this approximation by computing the exact $\rnni$ distance for small trees.
Specifically, we will use the algorithm for computing the complete $\rnni$ graph that is given in \autocite[their Section 3.3]{Gavryushkin2018-ol}.
After pre-computing the graph, we use the standard algorithms~\autocite{Floyd1962-ew,Dijkstra1959-ph} for computing shortest paths.
However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances for larger trees.
Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we compute the $\rnni$ graph on up to six taxa for our analyses.
This computational result will also be used later in the paper to support Conjecture~\ref{conjecture:cluster_theorem}, a modification of Conjecture~9 in \autocite{Gavryushkin2018-ol}, which we prove to be false in this paper.

%TODO: Make a zenodo dio for the repository before final publication
We have implemented all these algorithms in Python and the source code is available at \autocite{Collienne2019}.

@@

The algorithm $\csort$ (Algorithm~\ref{alg:csort}) is a natural modification of the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi}.
A path $p$ from $T$ to $R= [x_1, x_2, \ldots, x_n]$ is computed iteratively so that after $k$ steps the last $k$ taxa of $T$ and $R$ coincide.
Specifically, in step $k$ ($k = 0, \ldots, n-3$) the parent of taxon $x_{n-k}$ is moved up by $\nni$ moves to the position it has in $R$.
Notice that there might be more than just one such move per step, which means that $p$ can be extended by more than one tree in each step.
After $n-2$ steps we reach $R$.

The path $p$ returned by the algorithm $\csort$ is a shortest caterpillar path because every tree modification on $p$ reduces the number of inversions of taxa in $T$ and $R$.
Our Python implementation of this algorithm can be accessed at \autocite{Collienne2019}.
